---
title: "Clasper length analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clasper length analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(ggplot2)
set.seed(42)
```

## Background

Male chondrichthyans undergo a rapid elongation and calcification of the
claspers as they mature. Clasper length plotted against body length or age is
therefore sigmoid: near-constant in juveniles, steeply increasing through
maturation, and asymptotic in adults. `clasp_length()` fits that curve:

$$CL = b + (a - b)\left(1 + e^{-\log(19)\,(x - L_{50})/(L_{95} - L_{50})}\right)^{-1}$$

where $b$ and $a$ are the juvenile and adult asymptotes, and $L_{50}$ and
$L_{95}$ are the values of the predictor at which clasper length reaches 50% and
95% of the way between them. The $\log(19)$ term is what makes $L_{50}$ and
$L_{95}$ mean exactly that — the same reparameterisation used by `maturity()`,
so the two are directly comparable.

The model is fitted on the log scale by `nls()`. This gives multiplicative
error, matching the way variability in clasper length grows with clasper size,
and means the fitted curve describes *median* clasper length at a given size.

## Basic usage

Females have no clasper measurement and are dropped automatically, so the whole
data frame can be passed in:

```{r}
library(mustelus)
data(spottail)

cl <- clasp_length(clasp_length, length, clasp_calc, data = spottail, times = 200)
summary(cl)
```

Claspers are around `r round(summary(cl)$b)` mm in juveniles and asymptote near
`r round(summary(cl)$a)` mm in adults, with the transition centred on
`r round(summary(cl)$L50)` mm.

## Plotting

```{r}
plot(cl) + xlab("Total length (mm)") + ylab("Clasper length (mm)")
```

The solid line is the fitted median, the dashed ribbon the 95% confidence
interval and the dotted ribbon the 95% prediction interval. Both are obtained by
bootstrap: the confidence interval from the percentiles of the resampled curves,
and the prediction interval by combining that bootstrap variance with the
residual variance of the fit.

Passing the optional calcification variable shades the points by stage — white
for uncalcified, grey for partially calcified and black for fully calcified.
This is a useful check on the fitted curve, since the three stages should
separate along it: uncalcified animals on the lower asymptote, partially
calcified animals through the steep transition, and calcified animals on the
upper asymptote. Calcification is used only for plotting and does not enter the
model.

## Using age as the predictor

```{r}
cl_age <- clasp_length(clasp_length, age_agree, clasp_calc, data = spottail, times = 200)
summary(cl_age)
```

```{r}
plot(cl_age) + xlab("Age (years)") + ylab("Clasper length (mm)")
```

## Starting values

`nls()` needs starting values, and `clasp_length()` derives them from the data:
the asymptotes from the mean clasper length of the smallest and largest 20% of
animals, and $L_{50}$ and $L_{95}$ by interpolating binned means. Fitting uses
the `"port"` algorithm with both asymptotes constrained positive, since a
negative asymptote would make the log-scale model undefined.

If a fit fails to converge, supply starting values directly:

```{r}
clasp_length(clasp_length, length,
  data = spottail, times = 50,
  start = list(a = 80, b = 13, L50 = 900, L95 = 990)
) |> summary()
```

## Comparison with maturity

Because both functions report $L_{50}$ on the same scale, clasper elongation can
be compared directly with the onset of maturity:

```{r, warning = FALSE, message = FALSE}
mat <- maturity(maturity_stage, length, data = spottail[spottail$sex == "m", ],
                times = 200)

data.frame(
  quantity = c("Clasper elongation", "Maturity"),
  L50      = c(summary(cl)$L50, summary(mat)$L50),
  L95      = c(summary(cl)$L95, summary(mat)$L95)
)
```

## Output structure

The result is a one row tibble with list columns, including the bootstrap
parameter estimates:

```{r}
# Fitted nls object
cl$mods[[1]]
```

```{r}
# Bootstrap parameter estimates, one row per replicate
head(cl$boot_coefs[[1]])
```
