---
title: "Growth analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Growth analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
library(ggplot2)
```

## Background

Growth is usually described with a von Bertalanffy function, and usually
parameterised by $L_\infty$, $K$ and $t_0$. The third parameter is the
hypothetical age at zero length, which has no biological meaning and is
often estimated at implausible values when there are few young animals in
the sample. `growth()` uses the reparameterisation of Harry et al. (2019),
replacing $t_0$ with length at birth:

$$L_{i,g}(a) = L_0 + (L_{\infty,g} - L_0)\left(1 - e^{-K_g a_{i,g}}\right)$$

$L_0$ is interpretable, can be measured directly, and anchors the curve at
the left-hand end where age data are usually weakest. $L_\infty$ and $K$
are estimated separately for each group, typically sex, while $L_0$ is
shared.

Observed length is normally distributed about the curve with a standard
deviation proportional to expected length, $\sigma_L = CV_L L_{i,g}$,
following Cope and Punt (2007). Individual variability in length at age
therefore grows with size, which is what length at age data normally look
like.

Two further components from Harry et al. (2019) are optional, and can be
used independently.

### Ageing error

Ages read from vertebrae are estimates, not measurements. Treating them as
exact can bias growth estimates, by an amount that depends on how
imprecise the readings are and where in the age range the error falls.
The model treats true age as a random effect, with each reading an
observation of it:

$$A_{i,j} = a_i + \epsilon_{a,ij}, \qquad \epsilon_{a,ij} \sim N(0, (CV_a a_i)^2)$$

The ageing CV is computed outside the likelihood from replicate readings,
following Chang (1982), and held fixed.

### Neonates

Animals with an unhealed umbilical scar are known to be age zero. Their
lengths are direct information on $L_0$:

$$L_{0,i} \sim N(L_0, (CV_L L_{0,i})^2)$$

These are often available in numbers, and from animals that were never
aged, so they add information at no cost.

Fitting is by maximum likelihood with `RTMB`, the random effects
integrated out by the Laplace approximation. Confidence intervals on the
curve come from the delta method; prediction intervals add $CV_L$.

This is a port of the `TMB` implementation used in Harry et al. (2019).
The examples below use the data from that study.

## Data

```{r}
library(mustelus)
data(blacktip)
```

`blacktip` contains `r nrow(blacktip)` common blacktip sharks,
*Carcharhinus limbatus*, from south-east Queensland and northern New South
Wales. Vertebrae were read by two readers for
`r sum(!is.na(blacktip$age_agree))` sharks, all of them from New South
Wales, where most sharks were caught on demersal setlines. A further
`r sum(blacktip$neonate)` neonates with open or partially healed umbilical
scars were sampled in Moreton Bay, Queensland. None of the neonates were
aged. Length is stretched total length in cm.

## Basic usage

```{r}
g <- growth(STL, age_agree, sex, data = blacktip, neonate = neonate)
summary(g)
```

The summary has one row per group. $L_\infty$ and $K$ differ between them;
$L_0$ and $CV_L$ are shared and so repeat. Also reported are the number of
aged animals per group (`n`), the number of neonates (`n0`), the ageing CV
if used, the negative log-likelihood, AIC, and whether the fit converged
with a positive-definite Hessian.

Female blacktip sharks reach a larger asymptotic length than males
(`r round(summary(g)$Linf[1])` against `r round(summary(g)$Linf[2])` cm)
and grow more slowly toward it (`r round(summary(g)$K[1], 2)` against
`r round(summary(g)$K[2], 2)`). That pattern, females larger and slower,
is common in carcharhinids.

## Plotting

```{r}
plot(g) + xlab("Age (years)") + ylab("Stretched total length (cm)")
```

The solid line is the fitted curve, the dashed ribbon the 95% confidence
interval on it, and the dotted ribbon the 95% prediction interval for
individual animals. The prediction interval widens with size, which is the
$CV_L$ assumption at work.

## Comparing growth between sexes

```{r, include = FALSE}
V <- g$mods[[1]]$sdreport$cov.fixed
r_lk <- sprintf("%.2f", cov2cor(V)[cbind(which(rownames(V) == "Linf"),
                                         which(rownames(V) == "K"))])
k_ci <- sprintf("%.3f", c(summary(g)$K_lower, summary(g)$K_upper))
```

$L_\infty$ and $K$ are estimated from the same data and are strongly
correlated, `r r_lk[1]` for females and `r r_lk[2]` for males. A larger
asymptote can be traded against slower growth toward it with little change
in the fit. The confidence intervals reported for each parameter
separately are therefore not a good basis for comparing groups, and a
joint confidence region showing the combinations of the two that are
consistent with the data is more informative.

Kimura (1980) constructed these regions from contours of the residual sum
of squares, $S$, which for normally distributed errors are contours of
equal likelihood. A combination of parameters lies inside the approximate
95% region if

$$S(L_\infty, K, t_0) \le \hat{S}\left(1 + \frac{p}{N - p} F_{p, N-p}(0.95)\right)$$

where $\hat{S}$ is the minimum, $N$ is the number of observations and
$p = 3$ is the number of parameters. A region in three dimensions is hard
to display, so Kimura conditioned on $t_0$ at its estimate and drew the
two-dimensional cross-section for $L_\infty$ and $K$.

`growth()` maximises a likelihood rather than minimising a sum of squares,
because the standard deviation changes with length. The same criterion in
terms of the log-likelihood $\ell$ is

$$2\left[\hat{\ell} - \ell(L_\infty, K)\right] \le N \log\left(1 + \frac{p}{N - p} F_{p, N-p}(0.95)\right)$$

which for a single group with constant variance is identical to Kimura's.
`growth_region()` applies this criterion to a `growth()` fit. $L_0$ takes
the place of $t_0$ and is held at its estimate, and $CV_L$ is re-estimated
at each point, as $\sigma$ effectively is in the ratio of sums of squares.
$N$ is the number of aged sharks of that sex. As in Kimura (1980), the
boundary is found by stepping through values of $K$ and solving for the two
values of $L_\infty$ at which the likelihood reaches the critical value.

```{r}
r <- growth_region(g)
summary(r)
```

`summary()` gives the estimates and the range of each parameter within the
region.

```{r}
plot(r) + labs(x = "Linf (cm)", y = "K (per year)", linetype = "Sex")
```

```{r, include = FALSE}
sr_f <- summary(r)[1, ]
above <- round(sr_f$Linf_max - sr_f$Linf)
below <- round(sr_f$Linf - sr_f$Linf_min)
```

The regions are not symmetrical about the estimates, as an ellipse drawn
from the covariance matrix would be. Each follows the curved ridge in the
likelihood surface along which $L_\infty$ is traded against $K$, and
extends further above the estimates than below them. For females the
region reaches about `r above` cm above the estimate of $L_\infty$ but
only `r below` cm below it. The intervals on $K$ alone overlap between the
sexes (`r k_ci[1]` to `r k_ci[3]` for females, `r k_ci[2]` to `r k_ci[4]`
for males), but the two regions do not. Taken together, the two
parameters describe clearly different growth curves for females and
males.

As Kimura noted for $t_0$, holding $L_0$ fixed means these are
cross-sections rather than true confidence regions, since more extreme
values of $L_\infty$ and $K$ may occur at a different $L_0$. Setting
`profile_L0 = TRUE` re-estimates $L_0$ at each point as well, giving the
projection of the full three-parameter region. Here that makes almost no
difference, because $L_0$ is closely determined by the neonates. Without
neonates or young animals $L_0$ is poorly determined, and the projection
can be noticeably wider, mainly in $K$.

`growth_region()` also works on models fitted with ageing error. Each
evaluation of the likelihood then includes the Laplace approximation, so
the calculation takes noticeably longer.

## Length at birth

Supplying `neonate` flags animals of known age zero. Any flagged animal
that also carries an age is used only for the length at birth term, not as
length at age data: an age-zero observation and a length at birth
observation say exactly the same thing, so counting both would double the
weight. The function reports how many were moved.

In `blacktip` the neonates come from a separate sample, and the aged sharks
include only `r sum(g$data[[1]]$age < 2)` under two years old. Fitting the
model without the neonates shows what they contribute:

```{r}
g_no_neo <- growth(STL, age_agree, sex, data = blacktip)

rbind(
  with_neonates    = summary(g)[1, c("L0", "L0_lower", "L0_upper")],
  without_neonates = summary(g_no_neo)[1, c("L0", "L0_lower", "L0_upper")]
)
```

The estimate of $L_0$ hardly changes, but the confidence interval is about
five times narrower with the neonates included. Neonates are frequently
the only data available for the smallest sizes.

## Ageing error

Where replicate readings exist, pass the columns to `reads` and the ageing
CV is computed from them. The readings need to be on the same scale as
`age`. A reading is a count of growth zones and so is a whole number, but
the ages in `blacktip` also include a fractional adjustment for the time
elapsed since the population birth date of 1 November. That adjustment
depends on the date of capture rather than on the reader, so it is the
same for every reading of a given animal and is added to each one.

```{r}
library(dplyr)

aged <- filter(blacktip, !is.na(age_agree))
table(difference = round(aged$reader2 - aged$reader1))
```

The two readers agreed on `r sum(aged$reader1 == aged$reader2)` of the
`r nrow(aged)` sharks. Most disagreements were of a single zone, and
disagreement was far more common in sharks older than ten years. This is
the usual pattern for vertebral ageing.

```{r}
g_reads <- growth(STL, age_agree, sex, data = blacktip, neonate = neonate,
                  reads = c(reader1, reader2))
summary(g_reads)[, c("group", "Linf", "K", "L0", "cv_age")]
```

```{r, include = FALSE}
s0 <- summary(g)
sr <- summary(g_reads)
k_change <- round(100 * (sr$K - s0$K) / s0$K, 1)
linf_change <- round(abs(sr$Linf - s0$Linf), 1)
```

The ageing CV is `r round(100 * sr$cv_age[1], 1)`%. Compared with the fit
that treats age as exact, $K$ falls by `r abs(k_change[1])`% for females
and `r abs(k_change[2])`% for males, and $L_\infty$ moves by less than
`r ceiling(max(linf_change))` cm. Both shifts are well inside the
confidence intervals. Most of the disagreement between readers is in older
sharks, where the curve is nearly flat and length says little about which
reading is closer to the truth. The model treats the ages of those sharks
as more uncertain, but does not move them much. With less precise ageing
the effect can be considerably larger.

These estimates differ slightly from those published in Harry et al.
(2019). The loop over readings in the original code started at the second
reader, so the first reader's counts were never used. Fitted to the second
reader alone, `growth()` reproduces the published estimates to within
rounding: $L_{\infty}$ 263.3 and 241.9 cm, $K$ 0.1418 and 0.1565, $L_0$
72.77 cm and $CV_L$ 0.0487.

Leaving the birth date adjustment off the readings is an easy mistake to
make, and it is not a harmless one:

```{r}
no_adj <- blacktip |>
  mutate(reader1 = floor(reader1), reader2 = floor(reader2))

g_no_adj <- growth(STL, age_agree, sex, data = no_adj, neonate = neonate,
                   reads = c(reader1, reader2))
summary(g_no_adj)[, c("group", "Linf", "K", "L0", "cv_age")]
```

The readings now sit around half a year below the true ages, and the
model concludes that every shark is younger than it is. Because the
neonates hold $L_0$ in place, the error goes into the rest of the curve
instead: for females $K$ rises from `r round(summary(g_reads)$K[1], 3)` to
`r round(summary(g_no_adj)$K[1], 3)`, and $L_\infty$ falls by
`r round(summary(g_reads)$Linf[1] - summary(g_no_adj)$Linf[1])` cm.

If only a consensus age is available but the ageing CV is known from
elsewhere, supply it directly with `cv_age`. This is the more common
situation when reanalysing published data.

```{r}
g_cv <- growth(STL, age_agree, sex, data = blacktip, neonate = neonate,
               cv_age = 0.08)
summary(g_cv)[, c("group", "Linf", "K", "L0", "cv_age")]
```

$K$ falls, as it did with the two readings, though by less. The two
analyses are not equivalent, because the data going in are not the same.
The consensus age matches the first reader's count for
`r sum(aged$age_agree == aged$reader1)` of the `r nrow(aged)` sharks, and
differs from the mean of the two readings for
`r sum(abs((aged$reader1 + aged$reader2) / 2 - aged$age_agree) > 1e-9)`.

With neither `reads` nor `cv_age`, age is treated as measured without
error and the model reduces to an ordinary von Bertalanffy fit.

## Output structure

```{r}
# Fixed-effect estimates and standard errors
summary(g$mods[[1]]$sdreport, "fixed")
```

```{r}
head(g$preds[[1]])
```

`mods` holds the `RTMB` objective function, the `nlminb` result and the
`sdreport`. Note that the objective is an external pointer and will not
survive `saveRDS()`; the rest of the object is ordinary R data.

## References

Chang, W.Y.B. (1982) A statistical method for evaluating the
reproducibility of age determination. *Canadian Journal of Fisheries and
Aquatic Sciences* **39**(8), 1208–1210.
[doi:10.1139/f82-158](https://doi.org/10.1139/f82-158)

Cope, J.M. and Punt, A.E. (2007) Admitting ageing error when fitting
growth curves: an example using the von Bertalanffy growth function with
random effects. *Canadian Journal of Fisheries and Aquatic Sciences*
**64**(2), 205–218.
[doi:10.1139/f06-179](https://doi.org/10.1139/f06-179)

Harry, A.V., Butcher, P.A., Macbeth, W.G., Morgan, J.A.T., Taylor, S.M.
and Geraghty, P.T. (2019) Life history of the common blacktip shark,
*Carcharhinus limbatus*, from central eastern Australia and comparative
demography of a cryptic shark complex. *Marine and Freshwater Research*
**70**(6), 834–848.
[doi:10.1071/MF18141](https://doi.org/10.1071/MF18141)

Kimura, D.K. (1980) Likelihood methods for the von Bertalanffy growth
curve. *Fishery Bulletin* **77**(4), 765–776.
