---
title: "Growth analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Growth analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
library(ggplot2)

# Figures are numbered in the order listed in figs. fig() gives a linked
# reference for the text and fig_cap() a numbered caption. Each captioned
# figure is anchored by its chunk label, which is the link target
figs <- c("vb-rate", "vb-curve", "growth-fit", "growth-region")
fig <- function(id) sprintf("[Figure %d](#%s)", match(id, figs), id)
fig_cap <- function(text) {
  sprintf("**Figure %d.** %s", match(knitr::opts_current$get("label"), figs), text)
}
knitr::knit_hooks$set(fig.cap = function(before, options) {
  if (before) sprintf('\n\n<div id="%s"></div>\n\n', options$label)
})
```

```{css, echo = FALSE}
/* Set figure captions apart from the body text. The first selector is used
   by rmarkdown::html_vignette and the second by pkgdown */
.figcaption, p.caption {
  font-size: 0.9em;
  color: #555;
  margin-top: 0.25em;
  margin-bottom: 1.5em;
}
```

## Background

Growth in fishes is usually described with the von Bertalanffy growth
function (von Bertalanffy 1938). Von Bertalanffy derived it from the
balance between the synthesis and breakdown of body tissue, and in the
form used for length it assumes that growth rate declines linearly as
length increases:

$$\frac{dL}{dt} = K(L_\infty - L)$$

Here $L$ is length and $dL/dt$ is growth rate, the change in length
over time, $t$. Growth is fastest in the smallest animals and slows as
they approach $L_\infty$, the asymptotic length, at which growth rate
reaches zero (`r fig("vb-rate")`). $L_\infty$ is usually interpreted as the
average maximum length of individuals in the population. $K$ is the
growth coefficient, with units of $\text{year}^{-1}$, and is the slope
of the relationship between growth rate and length. It is not itself a
growth rate. Rather, it describes how quickly length approaches the
asymptote: an animal takes $\ln(2)/K$ years to grow half of the
remaining distance to $L_\infty$, whatever its current length.

```{r vb-rate, echo = FALSE, fig.height = 3.5, fig.cap = fig_cap("Growth rate, $dL/dt$, as a function of length, $L$, for values similar to those estimated for female blacktip sharks below ($L_0$ = 73 cm, $L_\\infty$ = 265 cm, $K$ = 0.14). The dotted line extends the relationship below length at birth.")}
L0 <- 73
Linf <- 265
K <- 0.14
t0 <- log(1 - L0 / Linf) / K

rate <- data.frame(L = seq(0, Linf, length.out = 200))
rate$dLdt <- K * (Linf - rate$L)

ggplot(rate, aes(L, dLdt)) +
  geom_line(data = rate[rate$L >= L0, ]) +
  geom_line(data = rate[rate$L <= L0, ], linetype = "dotted") +
  geom_vline(xintercept = c(L0, Linf), linetype = "dashed") +
  annotate("text", x = c(L0, Linf), y = K * Linf * 1.08,
           label = c("italic(L)[0]", "italic(L)['∞']"), parse = TRUE, hjust = -0.2) +
  annotate("text", x = (L0 + Linf) / 2, y = K * (Linf - (L0 + Linf) / 2),
           label = "slope == -italic(K)", parse = TRUE, hjust = -0.2, vjust = -0.6) +
  scale_y_continuous(limits = c(0, K * Linf * 1.12)) +
  labs(x = expression(italic(L)), y = expression(italic(dL/dt))) +
  theme_classic()
```

The solution to this equation most often used in fisheries science,
popularised by Beverton and Holt (1957), gives length as a function of
age, $a$:

$$L(a) = L_\infty\left(1 - e^{-K(a - t_0)}\right)$$

The third parameter, $t_0$, is the hypothetical age at which length
would be zero. Many teleost fishes hatch as small larvae, and for these
$t_0$ has no biological meaning. Chondrichthyans are born live or hatch
from egg cases at a relatively large size, and for them length at age
zero is meaningful: it is length at birth. The curve can be recast with
length at birth, $L_0$, the y-intercept, in place of $t_0$, the
x-intercept (`r fig("vb-curve")`):

$$L(a) = L_0 + (L_\infty - L_0)\left(1 - e^{-Ka}\right)$$

The two forms give identical lengths at age, and either intercept can be
calculated from the other, since $L_0 = L_\infty(1 - e^{K t_0})$ (Harry
et al. 2022). For the curve in `r fig("vb-curve")`, a length at birth of 73 cm
corresponds to a $t_0$ of `r round(t0, 1)` years, a little over two
years before birth. Holden (1974) suggested that $-t_0$ could be taken
as the gestation period, on the assumption that embryos grow along the
same curve as animals after birth, which gave a way of estimating $K$
for species that had not been aged (e.g. Francis 1981). Later
comparisons with $K$ from age and growth studies showed little
agreement, and the assumption about embryonic growth is unsupported for
most species (Pratt and Casey 1990).

```{r vb-curve, echo = FALSE, fig.height = 3.5, fig.cap = fig_cap(sprintf("The relationship in %s expressed as length at age. The first form of the equation defines the curve by where it would cross zero length, $t_0$; the second by where it starts, at length at birth, $L_0$.", fig("vb-rate")))}
len_age <- data.frame(a = seq(t0, 25, length.out = 300))
len_age$L <- Linf * (1 - exp(-K * (len_age$a - t0)))

ggplot(len_age, aes(a, L)) +
  geom_hline(yintercept = Linf, linetype = "dashed") +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_line(data = len_age[len_age$a >= 0, ]) +
  geom_line(data = len_age[len_age$a <= 0, ], linetype = "dotted") +
  geom_point(data = data.frame(a = c(t0, 0), L = c(0, L0))) +
  annotate("text", x = c(t0, 0, 25), y = c(0, L0, Linf),
           label = c("italic(t)[0]", "italic(L)[0]", "italic(L)['∞']"), parse = TRUE,
           hjust = c(-0.4, -0.4, 1), vjust = c(-0.6, 1.4, -0.6)) +
  scale_y_continuous(limits = c(0, Linf * 1.1)) +
  labs(x = "Age (years)", y = "Length (cm)") +
  theme_classic()
```

`growth()` uses the second form, as in Harry et al. (2019). $L_\infty$
and $K$ are estimated separately for each group, typically sex, while
$L_0$ is shared:

$$L_g(a) = L_0 + (L_{\infty,g} - L_0)\left(1 - e^{-K_g a}\right)$$

$L_0$ can be measured directly from neonates, and anchors the curve at
the left-hand end, where age data are usually weakest.

The equations above are deterministic. They describe the average length
at age, not the lengths of individual animals, and fitting them to data
requires some model of how individuals vary about the average. Growth
curves are commonly fitted by nonlinear least squares regression of
length on age (Harry et al. 2022). This implicitly assumes that the
observed length of animal $i$ is normally distributed about the curve
with a constant standard deviation, $\sigma$:

$$l_i = L(a_i) + \epsilon_i, \qquad \epsilon_i \sim N(0, \sigma^2)$$

Under this assumption a newborn and a 20 year old animal are equally
likely to differ from the average length for their age by any given
amount. In practice length at age usually becomes more variable as
animals get older and larger. A more realistic, but still simple,
alternative is for the coefficient of variation to be constant, so that
the standard deviation is proportional to expected length (Cope and Punt
2007; Restrepo et al. 2010):

$$\epsilon_i \sim N\left(0, (CV_L L(a_i))^2\right)$$

Restrepo et al. (2010) arrived at this form by allowing asymptotic
length to vary among individuals, following Kirkwood and Somers (1984).
It is the form used by `growth()`, with a single $CV_L$ shared between
groups and estimated along with the growth parameters.

Two further components from Harry et al. (2019) are optional, and can be
used independently.

### Ageing error

Ages read from vertebrae are estimates, not measurements. Treating them
as exact can bias growth estimates, by an amount that depends on how
imprecise the readings are and where in the age range the error falls.
The model treats true age as a random effect, with each reading an
observation of it:

$$A_{i,j} = a_i + \epsilon_{a,ij}, \qquad \epsilon_{a,ij} \sim N(0, (CV_a a_i)^2)$$

The ageing CV is computed outside the likelihood from replicate
readings, following Chang (1982), and held fixed.

### Neonates

Animals with an unhealed umbilical scar are known to be age zero. Their
lengths are direct information on $L_0$:

$$L_{0,i} \sim N(L_0, (CV_L L_{0,i})^2)$$

These are often available in numbers, and from animals that were never
aged, so they add information at no cost.

Fitting is by maximum likelihood with `RTMB`, the random effects
integrated out by the Laplace approximation. Confidence intervals on the
curve come from the delta method; prediction intervals add $CV_L$.

This is a port of the `TMB` implementation used in Harry et al. (2019).
The examples below use the data from that study.

## Data

```{r}
library(mustelus)
data(blacktip)
```

`blacktip` contains `r nrow(blacktip)` common blacktip sharks,
*Carcharhinus limbatus*, from south-east Queensland and northern New
South Wales. Vertebrae were read by two readers for
`r sum(!is.na(blacktip$age_agree))` sharks, all of them from New South
Wales, where most sharks were caught on demersal setlines. A further
`r sum(blacktip$neonate)` neonates with open or partially healed
umbilical scars were sampled in Moreton Bay, Queensland. None of the
neonates were aged. Length is stretched total length in cm.

## Basic usage

```{r}
g <- growth(STL, age_agree, sex, data = blacktip, neonate = neonate)
summary(g)
```

The summary has one row per group. $L_\infty$ and $K$ differ between
them; $L_0$ and $CV_L$ are shared and so repeat. Also reported are the
number of aged animals per group (`n`), the number of neonates (`n0`),
the ageing CV if used, the negative log-likelihood, AIC, and whether the
fit converged with a positive-definite Hessian.

Female blacktip sharks reach a larger asymptotic length than males
(`r round(summary(g)$Linf[1])` against `r round(summary(g)$Linf[2])` cm)
and grow more slowly toward it (`r round(summary(g)$K[1], 2)` against
`r round(summary(g)$K[2], 2)`). That pattern, females larger and slower,
is common in carcharhinids.

## Plotting

```{r growth-fit, fig.cap = fig_cap("Length at age of female (f) and male (m) blacktip sharks with the fitted growth curves. The solid line is the fitted curve, the dashed ribbon the 95% confidence interval and the dotted ribbon the 95% prediction interval. Neonates, which inform $L_0$, are not shown.")}
plot(g) + xlab("Age (years)") + ylab("Stretched total length (cm)")
```

In `r fig("growth-fit")` the solid line is the fitted curve, the dashed ribbon the 95% confidence
interval on it, and the dotted ribbon the 95% prediction interval for
individual animals. The prediction interval widens with size, which is
the $CV_L$ assumption at work.

## Comparing growth between sexes

```{r, include = FALSE}
V <- g$mods[[1]]$sdreport$cov.fixed
r_lk <- sprintf("%.2f", cov2cor(V)[cbind(which(rownames(V) == "Linf"),
                                         which(rownames(V) == "K"))])
k_ci <- sprintf("%.3f", c(summary(g)$K_lower, summary(g)$K_upper))
```

$L_\infty$ and $K$ are estimated from the same data and are strongly
correlated, `r r_lk[1]` for females and `r r_lk[2]` for males. A larger
asymptote can be traded against slower growth toward it with little
change in the fit. The confidence intervals reported for each parameter
separately are therefore not a good basis for comparing groups, and a
joint confidence region showing the combinations of the two that are
consistent with the data is more informative.

Kimura (1980) constructed these regions from contours of the residual
sum of squares, $S$, which for normally distributed errors are contours
of equal likelihood. A combination of parameters lies inside the
approximate 95% region if

$$S(L_\infty, K, t_0) \le \hat{S}\left(1 + \frac{p}{N - p} F_{p, N-p}(0.95)\right)$$

where $\hat{S}$ is the minimum, $N$ is the number of observations and
$p = 3$ is the number of parameters. A region in three dimensions is
hard to display, so Kimura conditioned on $t_0$ at its estimate and drew
the two-dimensional cross-section for $L_\infty$ and $K$.

`growth()` maximises a likelihood rather than minimising a sum of
squares, because the standard deviation changes with length. The same
criterion in terms of the log-likelihood $\ell$ is

$$2\left[\hat{\ell} - \ell(L_\infty, K)\right] \le N \log\left(1 + \frac{p}{N - p} F_{p, N-p}(0.95)\right)$$

which for a single group with constant variance is identical to
Kimura's. `growth_region()` applies this criterion to a `growth()` fit.
$L_0$ takes the place of $t_0$ and is held at its estimate, and $CV_L$
is re-estimated at each point, as $\sigma$ effectively is in the ratio
of sums of squares. $N$ is the number of aged sharks of that sex. As in
Kimura (1980), the boundary is found by stepping through values of $K$
and solving for the two values of $L_\infty$ at which the likelihood
reaches the critical value.

```{r}
r <- growth_region(g)
summary(r)
```

`summary()` gives the estimates and the range of each parameter within
the region.

```{r growth-region, fig.cap = fig_cap("Approximate 95% joint confidence regions for $L_\\infty$ and $K$ for female (f) and male (m) blacktip sharks, following Kimura (1980). Points are the maximum likelihood estimates.")}
plot(r) + labs(x = "Linf (cm)", y = "K (per year)", linetype = "Sex")
```

```{r, include = FALSE}
sr_f <- summary(r)[1, ]
above <- round(sr_f$Linf_max - sr_f$Linf)
below <- round(sr_f$Linf - sr_f$Linf_min)
```

The regions in `r fig("growth-region")` are not symmetrical about the estimates, as an ellipse drawn
from the covariance matrix would be. Each follows the curved ridge in
the likelihood surface along which $L_\infty$ is traded against $K$, and
extends further above the estimates than below them. For females the
region reaches about `r above` cm above the estimate of $L_\infty$ but
only `r below` cm below it. The intervals on $K$ alone overlap between
the sexes (`r k_ci[1]` to `r k_ci[3]` for females, `r k_ci[2]` to
`r k_ci[4]` for males), but the two regions do not. Taken together, the
two parameters describe clearly different growth curves for females and
males.

As Kimura noted for $t_0$, holding $L_0$ fixed means these are
cross-sections rather than true confidence regions, since more extreme
values of $L_\infty$ and $K$ may occur at a different $L_0$. Setting
`profile_L0 = TRUE` re-estimates $L_0$ at each point as well, giving the
projection of the full three-parameter region. Here that makes almost no
difference, because $L_0$ is closely determined by the neonates. Without
neonates or young animals $L_0$ is poorly determined, and the projection
can be noticeably wider, mainly in $K$.

`growth_region()` also works on models fitted with ageing error. Each
evaluation of the likelihood then includes the Laplace approximation, so
the calculation takes noticeably longer.

## Length at birth

Supplying `neonate` flags animals of known age zero. Any flagged animal
that also carries an age is used only for the length at birth term, not
as length at age data: an age-zero observation and a length at birth
observation say exactly the same thing, so counting both would double
the weight. The function reports how many were moved.

In `blacktip` the neonates come from a separate sample, and the aged
sharks include only `r sum(g$data[[1]]$age < 2)` under two years old.
Fitting the model without the neonates shows what they contribute:

```{r}
g_no_neo <- growth(STL, age_agree, sex, data = blacktip)

rbind(
  with_neonates    = summary(g)[1, c("L0", "L0_lower", "L0_upper")],
  without_neonates = summary(g_no_neo)[1, c("L0", "L0_lower", "L0_upper")]
)
```

The estimate of $L_0$ hardly changes, but the confidence interval is
about five times narrower with the neonates included. Neonates are
frequently the only data available for the smallest sizes.

## Ageing error

Where replicate readings exist, pass the columns to `reads` and the
ageing CV is computed from them. The readings need to be on the same
scale as `age`. A reading is a count of growth zones and so is a whole
number, but the ages in `blacktip` also include a fractional adjustment
for the time elapsed since the population birth date of 1 November. That
adjustment depends on the date of capture rather than on the reader, so
it is the same for every reading of a given animal and is added to each
one.

```{r}
library(dplyr)

aged <- filter(blacktip, !is.na(age_agree))
table(difference = round(aged$reader2 - aged$reader1))
```

The two readers agreed on `r sum(aged$reader1 == aged$reader2)` of the
`r nrow(aged)` sharks. Most disagreements were of a single zone, and
disagreement was far more common in sharks older than ten years. This is
the usual pattern for vertebral ageing.

```{r}
g_reads <- growth(STL, age_agree, sex, data = blacktip, neonate = neonate,
                  reads = c(reader1, reader2))
summary(g_reads)[, c("group", "Linf", "K", "L0", "cv_age")]
```

```{r, include = FALSE}
s0 <- summary(g)
sr <- summary(g_reads)
k_change <- round(100 * (sr$K - s0$K) / s0$K, 1)
linf_change <- round(abs(sr$Linf - s0$Linf), 1)
```

The ageing CV is `r round(100 * sr$cv_age[1], 1)`%. Compared with the
fit that treats age as exact, $K$ falls by `r abs(k_change[1])`% for
females and `r abs(k_change[2])`% for males, and $L_\infty$ moves by
less than `r ceiling(max(linf_change))` cm. Both shifts are well inside
the confidence intervals. Most of the disagreement between readers is in
older sharks, where the curve is nearly flat and length says little
about which reading is closer to the truth. The model treats the ages of
those sharks as more uncertain, but does not move them much. With less
precise ageing the effect can be considerably larger.

These estimates differ slightly from those published in Harry et al.
(2019). The loop over readings in the original code started at the
second reader, so the first reader's counts were never used. Fitted to
the second reader alone, `growth()` reproduces the published estimates
to within rounding: $L_{\infty}$ 263.3 and 241.9 cm, $K$ 0.1418 and
0.1565, $L_0$ 72.77 cm and $CV_L$ 0.0487.

Leaving the birth date adjustment off the readings is an easy mistake to
make. Here it gives the youngest sharks readings of zero, and `growth()`
stops:

```{r, error = TRUE}
no_adj <- blacktip |>
  mutate(reader1 = floor(reader1), reader2 = floor(reader2))

growth(STL, age_agree, sex, data = no_adj, neonate = neonate,
       reads = c(reader1, reader2))
```

Because the standard deviation of a reading is proportional to age, a
reading of zero can be explained exactly by a true age of zero, and the
likelihood has no maximum. Where no readings are zero the model will
fit, but the readings then sit below the true ages by the missing
fraction, around half a year on average here, and the model treats every
animal as younger than it is.

If only a consensus age is available but the ageing CV is known from
elsewhere, supply it directly with `cv_age`. This is the more common
situation when reanalysing published data.

```{r}
g_cv <- growth(STL, age_agree, sex, data = blacktip, neonate = neonate,
               cv_age = 0.08)
summary(g_cv)[, c("group", "Linf", "K", "L0", "cv_age")]
```

$K$ falls, as it did with the two readings, though by less. The two
analyses are not equivalent, because the data going in are not the same.
The consensus age matches the first reader's count for
`r sum(aged$age_agree == aged$reader1)` of the `r nrow(aged)` sharks,
and differs from the mean of the two readings for
`r sum(abs((aged$reader1 + aged$reader2) / 2 - aged$age_agree) > 1e-9)`.

With neither `reads` nor `cv_age`, age is treated as measured without
error and the model reduces to an ordinary von Bertalanffy fit.

## Output structure

```{r}
# Fixed-effect estimates and standard errors
summary(g$mods[[1]]$sdreport, "fixed")
```

```{r}
head(g$preds[[1]])
```

`mods` holds the `RTMB` objective function, the `nlminb` result and the
`sdreport`. Note that the objective is an external pointer and will not
survive `saveRDS()`; the rest of the object is ordinary R data.

## References

Beverton, R.J.H. and Holt, S.J. (1957) *On the Dynamics of Exploited
Fish Populations*. Fishery Investigations Series II, Volume 19. Ministry
of Agriculture, Fisheries and Food, London.

Chang, W.Y.B. (1982) A statistical method for evaluating the
reproducibility of age determination. *Canadian Journal of Fisheries and
Aquatic Sciences* **39**(8), 1208–1210.
[doi:10.1139/f82-158](https://doi.org/10.1139/f82-158)

Cope, J.M. and Punt, A.E. (2007) Admitting ageing error when fitting
growth curves: an example using the von Bertalanffy growth function with
random effects. *Canadian Journal of Fisheries and Aquatic Sciences*
**64**(2), 205–218.
[doi:10.1139/f06-179](https://doi.org/10.1139/f06-179)

Francis, M.P. (1981) Von Bertalanffy growth rates in species of
*Mustelus* (Elasmobranchii: Triakidae). *Copeia* **1981**(1), 189–192.
[doi:10.2307/1444053](https://doi.org/10.2307/1444053)

Harry, A.V., Butcher, P.A., Macbeth, W.G., Morgan, J.A.T., Taylor, S.M.
and Geraghty, P.T. (2019) Life history of the common blacktip shark,
*Carcharhinus limbatus*, from central eastern Australia and comparative
demography of a cryptic shark complex. *Marine and Freshwater Research*
**70**(6), 834–848.
[doi:10.1071/MF18141](https://doi.org/10.1071/MF18141)

Harry, A.V., Smart, J.J. and Pardo, S.A. (2022) Understanding the age
and growth of chondrichthyan fishes. In: Carrier, J.C., Simpfendorfer,
C.A., Heithaus, M.R. and Yopak, K.E. (eds) *Biology of Sharks and Their
Relatives*, 3rd edn. CRC Press, Boca Raton, FL, pp. 177–202.
[doi:10.1201/9781003262190-6](https://doi.org/10.1201/9781003262190-6)

Holden, M.J. (1974) Problems in the rational exploitation of
elasmobranch populations and some suggested solutions. In: Harden Jones,
F.R. (ed) *Sea Fisheries Research*. Halsted Press, John Wiley & Sons,
New York, pp. 117–137.

Kimura, D.K. (1980) Likelihood methods for the von Bertalanffy growth
curve. *Fishery Bulletin* **77**(4), 765–776.

Kirkwood, G.P. and Somers, I.F. (1984) Growth of two species of tiger
prawn, *Penaeus esculentus* and *P. semisulcatus*, in the western Gulf
of Carpentaria. *Australian Journal of Marine and Freshwater Research*
**35**(6), 703–712.
[doi:10.1071/MF9840703](https://doi.org/10.1071/MF9840703)

Pratt, H.L., Jr. and Casey, J.G. (1990) Shark reproductive strategies as
a limiting factor in directed fisheries, with a review of Holden's
method of estimating growth-parameters. In: Pratt, H.L., Jr., Gruber,
S.H. and Taniuchi, T. (eds) *Elasmobranchs as Living Resources: Advances
in the Biology, Ecology, Systematics, and the Status of the Fisheries*.
NOAA Technical Report NMFS 90, pp. 97–109.
[NOAA Technical Report NMFS 90](https://spo.nmfs.noaa.gov/content/tr-90-elasmobranchs-living-resources-advances-biology-ecology-systematics-and-status)

Restrepo, V.R., Diaz, G.A., Walter, J.F., Neilson, J.D., Campana, S.E.,
Secor, D. and Wingate, R.L. (2010) Updated estimate of the growth curve
of Western Atlantic bluefin tuna. *Aquatic Living Resources* **23**(4),
335–342.
[doi:10.1051/alr/2011004](https://doi.org/10.1051/alr/2011004)

von Bertalanffy, L. (1938) A quantitative theory of organic growth
(inquiries on growth laws. II). *Human Biology* **10**, 181–213.
