---
title: "Growth analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Growth analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
library(ggplot2)
set.seed(42)
```

## Background

Growth is usually described with a von Bertalanffy function, and usually
parameterised by $L_\infty$, $K$ and $t_0$. The third parameter is the
hypothetical age at zero length, which has no biological meaning and is
often estimated at implausible values when there are few young animals in
the sample. `growth()` uses the reparameterisation of Harry et al. (2019),
replacing $t_0$ with length at birth:

$$L_{i,g}(a) = L_0 + (L_{\infty,g} - L_0)\left(1 - e^{-K_g a_{i,g}}\right)$$

$L_0$ is interpretable, can be measured directly, and anchors the curve at
the left-hand end where age data are usually weakest. $L_\infty$ and $K$
are estimated separately for each group, typically sex, while $L_0$ is
shared.

Observed length is normally distributed about the curve with a standard
deviation proportional to expected length, $\sigma_L = CV_L L_{i,g}$,
following Cope and Punt (2007). Individual variability in length at age
therefore grows with size, which is what length at age data normally look
like.

Two further components from Harry et al. (2019) are optional, and can be
used independently.

### Ageing error

Ages read from vertebrae are estimates, not measurements. Treating them as
exact biases growth estimates: the usual consequence is that $K$ comes out
too high and $L_\infty$ too low, because reading error flattens the
apparent relationship at the top end. The model treats true age as a
random effect, with each reading an observation of it:

$$A_{i,j} = a_i + \epsilon_{a,ij}, \qquad \epsilon_{a,ij} \sim N(0, (CV_a a_i)^2)$$

The ageing CV is computed outside the likelihood from replicate readings,
following Chang (1982), and held fixed.

### Neonates

Animals with an unhealed umbilical scar are known to be age zero. Their
lengths are direct information on $L_0$:

$$L_{0,i} \sim N(L_0, (CV_L L_{0,i})^2)$$

These are often available in numbers, and from animals that were never
aged, so they add information at no cost.

Fitting is by maximum likelihood with `RTMB`, the random effects
integrated out by the Laplace approximation. Confidence intervals on the
curve come from the delta method; prediction intervals add $CV_L$.

This is a port of the `TMB` implementation used in Harry et al. (2019),
and reproduces the published estimates for *Carcharhinus limbatus* to
within rounding: $L_{\infty}$ 263.3 and 241.9 cm, $K$ 0.1418 and 0.1565,
$L_0$ 72.77 cm and $CV_L$ 0.0487.

## Basic usage

```{r}
library(mustelus)
data(spottail)

g <- growth(length, age_agree, sex, data = spottail,
            neonate = umb_scar %in% c("y", "p"))
summary(g)
```

The summary has one row per group. $L_\infty$ and $K$ differ between them;
$L_0$ and $CV_L$ are shared and so repeat. Also reported are the number of
aged animals per group (`n`), the number of neonates (`n0`), the ageing CV
if used, the negative log-likelihood, AIC, and whether the fit converged
with a positive-definite Hessian.

Female spot-tail sharks reach a larger asymptotic length than males
(`r round(summary(g)$Linf[1])` against `r round(summary(g)$Linf[2])` mm)
and grow more slowly toward it (`r round(summary(g)$K[1], 2)` against
`r round(summary(g)$K[2], 2)`). That pattern, females larger and slower,
is common in carcharhinids.

## Plotting

```{r}
plot(g) + xlab("Age (years)") + ylab("Total length (mm)")
```

The solid line is the fitted curve, the dashed ribbon the 95% confidence
interval on it, and the dotted ribbon the 95% prediction interval for
individual animals. The prediction interval widens with size, which is the
$CV_L$ assumption at work.

## Length at birth

Supplying `neonate` flags animals of known age zero. Any flagged animal
that also carries an age is used only for the length at birth term, not as
length at age data: an age-zero observation and a length at birth
observation say exactly the same thing, so counting both would double the
weight. The function reports how many were moved.

```{r}
g$neonates[[1]]
```

Here all five neonates were also aged, at exactly age zero, so the two
formulations are mathematically identical and removing them changes
nothing. Where neonates come from separate sampling, as in Harry et al.
(2019), they add genuinely new information, and are frequently the only
data available for the smallest sizes.

## Ageing error

Where replicate readings exist, pass the columns to `reads` and the ageing
CV is computed from them. `spottail` carries only a consensus age, so for
illustration here are two simulated readings around it:

```{r}
library(dplyr)

sp <- spottail |>
  mutate(reader1 = pmax(0, round(age_agree + rnorm(n(), 0, 0.3), 2)),
         reader2 = pmax(0, round(age_agree + rnorm(n(), 0, 0.3), 2)))

g_reads <- growth(length, age_agree, sex, data = sp,
                  neonate = umb_scar %in% c("y", "p"),
                  reads = c(reader1, reader2))
summary(g_reads)[, c("group", "Linf", "K", "L0", "cv_age")]
```

If only a consensus age is available but the ageing CV is known from
elsewhere, supply it directly with `cv_age`. This is the more common
situation when reanalysing published data.

```{r}
sapply(c(0, 0.05, 0.10, 0.20), function(cv) {
  gg <- if (cv == 0) {
    growth(length, age_agree, sex, data = spottail,
           neonate = umb_scar %in% c("y", "p"))
  } else {
    growth(length, age_agree, sex, data = spottail,
           neonate = umb_scar %in% c("y", "p"), cv_age = cv)
  }
  c(cv_age = cv, Linf_f = summary(gg)$Linf[1], K_f = summary(gg)$K[1])
}) |> t() |> as.data.frame()
```

As the assumed ageing error grows, $L_\infty$ rises and $K$ falls. That is
the expected direction: ignoring ageing error makes growth look faster and
the asymptote smaller than it is. The size of the shift depends on how much
error there is, which is why estimating $CV_a$ from replicate readings is
worth doing where the readings exist.

With neither `reads` nor `cv_age`, age is treated as measured without
error and the model reduces to an ordinary von Bertalanffy fit.

## Output structure

```{r}
# Fixed-effect estimates and standard errors
summary(g$mods[[1]]$sdreport, "fixed")
```

```{r}
head(g$preds[[1]])
```

`mods` holds the `RTMB` objective function, the `nlminb` result and the
`sdreport`. Note that the objective is an external pointer and will not
survive `saveRDS()`; the rest of the object is ordinary R data.

## References

Chang, W.Y.B. (1982) A statistical method for evaluating the
reproducibility of age determination. *Canadian Journal of Fisheries and
Aquatic Sciences* **39**(8), 1208–1210.
[doi:10.1139/f82-158](https://doi.org/10.1139/f82-158)

Cope, J.M. and Punt, A.E. (2007) Admitting ageing error when fitting
growth curves: an example using the von Bertalanffy growth function with
random effects. *Canadian Journal of Fisheries and Aquatic Sciences*
**64**(2), 205–218.
[doi:10.1139/f06-179](https://doi.org/10.1139/f06-179)

Harry, A.V., Butcher, P.A., Macbeth, W.G., Morgan, J.A.T., Taylor, S.M.
and Geraghty, P.T. (2019) Life history of the common blacktip shark,
*Carcharhinus limbatus*, from central eastern Australia and comparative
demography of a cryptic shark complex. *Marine and Freshwater Research*
**70**(6), 834–848.
[doi:10.1071/MF18141](https://doi.org/10.1071/MF18141)
