---
title: "Length-weight analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Length-weight analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(ggplot2)
```

## Background

The relationship between body mass and length in fishes is typically described by a power curve:

$$W_i = \alpha L_i^{\beta}$$

where $W_i$ and $L_i$ are the weight and length of individual $i$, and $\alpha$ and $\beta$ are estimated parameters. Individual variability in weight also increases with length, so a multiplicative rather than additive error structure is more appropriate:

$$W_i = \alpha L_i^{\beta} e^{\epsilon}, \quad \epsilon \sim N(0, \sigma^2)$$

Upon log-transformation this becomes a standard linear regression:

$$\ln(W_i) = \ln(\alpha) + \beta \cdot \ln(L_i) + \epsilon$$

where $\ln(\alpha)$ is the intercept and $\beta$ is the slope. This is the model fitted by `len_weight()`.

## Bias correction

A subtle but important consequence of log-transforming the response is that back-transforming predictions gives the *median* weight at a given length, not the mean. For a lognormal distribution, the mean is:

$$E[W \mid L] = e^{\hat{\mu} + \sigma^2/2}$$

where $\sigma^2$ is the residual variance. `len_weight()` applies this correction automatically — all predicted values and intervals are corrected for back-transformation bias.

## Basic usage

```{r}
library(mustelus)
data(spottail)

lw <- len_weight(wgt, length, data = spottail)
summary(lw)
```

The summary table reports: the intercept on the natural scale ($a = e^{\ln(\alpha)}$), its standard error range, the slope $b$, its standard error, the residual standard deviation $\sigma$, the log-likelihood, sample size, and length range.

## Grouping by sex

Pass a grouping variable to fit separate models per group. A pooled model across all groups is automatically included.

```{r}
lw_sex <- len_weight(wgt, length, sex, data = spottail)
summary(lw_sex)
```

## Plotting

`plot()` returns a named list of ggplot objects — one per group — which can be customised with standard ggplot2 calls. The solid line is the predicted mean, the dashed ribbon the 95% confidence interval, and the dotted ribbon the 95% prediction interval.

```{r}
p <- plot(lw_sex)
p$f + xlab("Total length (mm)") + ylab("Weight (kg)")
```

```{r}
p$m + xlab("Total length (mm)") + ylab("Weight (kg)")
```

## Confidence vs prediction intervals

Both interval types appear in the output. They answer different questions:

- **Confidence interval** — uncertainty in the *mean* weight at a given length. Narrows with more data.
- **Prediction interval** — expected spread of *individual* weights at a given length. Reflects the biological variability captured by $\sigma$ and does not narrow appreciably with more data.

For most fisheries reporting purposes the prediction interval is more informative, as it describes the range of weights you would expect to observe in the field.

## Using other length measurements

The function accepts any positive numeric predictor — fork length, precaudal length, or any other measurement that makes biological sense to log-transform.

```{r}
lw_fl <- len_weight(wgt, FL, data = spottail)
summary(lw_fl)
```

## Output structure

The result is a nested tibble with one row per group. List columns `data`, `coefs`, `preds`, and `mods` are named by group level for direct access:

```{r}
# Fitted lm object for females
lw_sex$mods[["f"]]
```

```{r}
# Predicted values for the pooled group
head(lw_sex$preds[["all"]])
```

The `preds` data frame contains six columns: `length`, `weight` (bias-corrected mean), `clower`/`cupper` (95% confidence interval), and `plower`/`pupper` (95% prediction interval).
