---
title: "Length-weight analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Length-weight analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(ggplot2)
```

The `len_weight()` function fits a log-linear regression of body mass as a function of length:

$$\log(W) = \log(a) + b \cdot \log(L)$$

which is equivalent to the standard fisheries length-weight relationship $W = aL^b$.

## Basic usage

```{r}
library(mustelus)
data(spottail)

lw <- len_weight(wgt, length, data = spottail)
summary(lw)
```

## Grouping by sex

Pass a grouping variable to fit separate models per group. A pooled model across all groups is automatically included.

```{r}
lw_sex <- len_weight(wgt, length, sex, data = spottail)
summary(lw_sex)
```

## Plotting

`plot()` returns a named list of ggplot objects — one per group — which can be customised with standard ggplot2 calls.

```{r}
p <- plot(lw_sex)
p$f + xlab("Total length (mm)") + ylab("Weight (kg)")
```

```{r}
p$m + xlab("Total length (mm)") + ylab("Weight (kg)")
```

## Using other length measurements

The function accepts any numeric predictor — fork length, precaudal length, age, or anything else that makes sense to log-transform.

```{r}
lw_fl <- len_weight(wgt, FL, data = spottail)
summary(lw_fl)
```

## Output structure

The result is a nested tibble with one row per group. List columns `data`, `coefs`, `preds`, and `mods` are named by group level for easy access:

```{r}
# Fitted lm object for females
lw_sex$mods[["f"]]

# Predicted values for the pooled group
head(lw_sex$preds[["all"]])
```
