---
title: "Length-weight analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Length-weight analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(ggplot2)
```

## Background

Length and weight are perhaps the two most routinely taken measurements
in fisheries science, although length is the far easier one to measure.
Length can be measured quickly and is more commonly the standard
measurement, especially for chondrichthyans. Weight is far more
challenging to collect, especially in the field. It needs a stable
platform and, if at sea, a compensating balance to accurately measure.
Weight also varies with gut fullness, body condition, and gonad state.
In the case of larger chondrichthyan species, it is often impractical to
measure adult animals that weight hundreds or thousands of kilograms.
For these reasons, good quality length weight relationships don't exist
for that many chondrichthyan species which is precisely why this
information is still worth collecting.

This conversion between length and weight underpins a considerable
amount of subsequent work in quantitative fisheries science. Catch is
reported and managed in tonnes whereas biological sampling is usually
done in numbers of fish at length, so moving between the two requires a
weight–length relationship, as does converting population length
structure (the distribution of lengths in a population) to biomass. The
age-structured population dynamics models that are commonly used in
stock assessments need weight-at-age, which is normally obtained by
passing predicted length-at-age from a growth curve through the
weight–length relationship; yield-per-recruit, spawning stock biomass
and other demographic calculations follow. The same relationship
underlies body condition indices, which compare the observed weight of
an individual against the weight expected for its length (Froese, 2006).

Despite its importance, the length weight relationship is considered
trivial, as noted by Froese (2006) who quotes *Quantitative Fisheries
Stock Assessment* (Hilborn and Walters, 2001): length–weight analysis is
"a good thing to have your teenage children do" as a way of learning
about correlation and regression, and if the teenager struggles with
estimating $\beta$, one can take comfort in assuming $\beta = 3$.

The relationship between body mass and length in fishes is typically
described by a power curve:

$$W_i = \alpha L_i^{\beta}$$

where $W_i$ and $L_i$ are the weight and length of individual $i$, and
$\alpha$ and $\beta$ are estimated parameters. Individual variability in
weight also increases with length, so a multiplicative rather than
additive error structure is more appropriate:

$$W_i = \alpha L_i^{\beta} e^{\epsilon}, \quad \epsilon \sim N(0, \sigma^2)$$

Upon log-transformation this becomes a standard linear regression:

$$\ln(W_i) = \ln(\alpha) + \beta \cdot \ln(L_i) + \epsilon$$

where $\ln(\alpha)$ is the intercept and $\beta$ is the slope. This is
the model fitted by `len_weight()`, which reports $\alpha$ as `a` and
$\beta$ as `b`.

## Interpreting the parameters

$\beta$, the exponent, is the more informative of the two parameters. It
describes how weight scales with length:

- $\beta = 3$ — **isometric** growth. Weight rises as the cube of
  length, so body shape and specific gravity stay constant as the animal
  grows and a fish that doubles in length becomes eight times heavier
  (i.e. the cube law).
- $\beta > 3$ — **positive allometry**. Large individuals are relatively
  heavier for their length, gaining girth or depth faster than length.
- $\beta < 3$ — **negative allometry**. Large individuals are relatively
  more slender for the length.

Froese's (2006) meta-analysis of 3929 relationships across 1773 species
found a median $\beta$ of 3.03, significantly greater than 3, so a
slight tendency toward positive allometry is the norm rather than the
exception. Froese (2006) indicates that $2.5 < \beta < 3.5$ is the
expected range for this coefficient.

$\alpha$, the coefficient, is the intercept on the log scale and
reflects body form: at a given $\beta$, eel-like and elongate species
carry smaller values than fusiform and deep-bodied species. It is also
important to note that $\alpha$ depends on the measurement units and on
which length measurement was used, changing by a factor of $10^{\beta}$
when length is converted from millimetres to centimetres. Values are
therefore not comparable across studies unless the units are the same.
Froese (2006) provides some of these commonly used conversion factors.

## Bias correction

A subtle consequence of log-transforming in the regression of weight
against length is that back-transforming predictions gives the *median*
weight at a given length, not the mean. For a lognormal distribution,
the mean is:

$$E[W \mid L] = e^{\hat{\mu} + \sigma^2/2}$$

where $\sigma^2$ is the residual variance. `len_weight()` accounts for
this — all predicted values and intervals are corrected for
back-transformation bias, following Beauchamp and Olson (1973).

The size of the correction is $e^{\sigma^2/2}$, so it depends only on
the residual variance and is constant across the length range. It is
negligible for a tight relationship and grows quickly once residual
variation is substantial: at $\sigma = 0.1$ it inflates predictions by
0.5%, at $\sigma = 0.3$ by 4.6%, and at $\sigma = 0.5$ by 13%. Reported
values of $\sigma$ therefore indicate how much difference the correction
makes for a given data set.

## Basic usage

```{r}
library(mustelus)
data(spottail)

lw <- len_weight(wgt, length, data = spottail)
summary(lw)
```

The summary table reports: the intercept on the natural scale
($a = e^{\ln(\alpha)}$), its standard error range, the slope $b$, its
standard error, the residual standard deviation $\sigma$, the
log-likelihood, sample size, and length range.

In the case of the spot-tail data the exponent,
`r round(summary(lw)$b, 2)`, is well above 3 and sits near the top of
Froese's expected range, indicating positive allometry: larger spottail
sharks are relatively heavier for their length than smaller ones. And
`a` is around `r format(summary(lw)$a, digits = 2)`. This value
corresponds to the median weight in kilograms of a one millimetre long
shark, extrapolated from a curve fitted to individuals two to three
orders of magnitude larger, so it doesn't have any particularly
important biological interpretation.

## Grouping by sex

Pass a grouping variable to fit separate models per group. A pooled
model across all groups is automatically included.

```{r}
lw_sex <- len_weight(wgt, length, sex, data = spottail)
summary(lw_sex)
```

## Plotting

`plot()` returns a named list of ggplot objects — one per group — which
can be customised with standard ggplot2 calls. The solid line is the
predicted mean, the dashed ribbon the 95% confidence interval, and the
dotted ribbon the 95% prediction interval.

```{r}
p <- plot(lw_sex)
p$f + xlab("Total length (mm)") + ylab("Weight (kg)")
```

```{r}
p$m + xlab("Total length (mm)") + ylab("Weight (kg)")
```

## Confidence vs prediction intervals

Both interval types appear in the output. They answer different
questions:

- **Confidence interval** — uncertainty in the *mean* weight at a given
  length. Narrows with more data.
- **Prediction interval** — expected spread of *individual* weights at a
  given length. Reflects the biological variability captured by $\sigma$
  and does not narrow appreciably with more data.

For most fisheries reporting purposes the prediction interval is more
informative, as it describes the range of weights you would expect to
observe in the field.

## Using other length measurements

The function accepts any positive numeric predictor — fork length,
precaudal length, or any other measurement that makes biological sense
to log-transform.

```{r}
lw_fl <- len_weight(wgt, FL, data = spottail)
summary(lw_fl)
```

## Output structure

The result is a nested tibble with one row per group. List columns
`data`, `coefs`, `preds`, and `mods` are named by group level for direct
access:

```{r}
# Fitted lm object for females
lw_sex$mods[["f"]]
```

```{r}
# Predicted values for the pooled group
head(lw_sex$preds[["all"]])
```

The `preds` data frame contains six columns: `length`, `weight`
(bias-corrected mean), `clower`/`cupper` (95% confidence interval), and
`plower`/`pupper` (95% prediction interval).

## References

Beauchamp, J.J. and Olson, J.S. (1973) Corrections for bias in
regression estimates after logarithmic transformation. *Ecology*
**54**(6), 1403–1407.
[doi:10.2307/1934208](https://doi.org/10.2307/1934208)

Froese, R. (2006) Cube law, condition factor and weight–length
relationships: history, meta-analysis and recommendations. *Journal of
Applied Ichthyology* **22**(4), 241–253.
[doi:10.1111/j.1439-0426.2006.00805.x](https://doi.org/10.1111/j.1439-0426.2006.00805.x)

Hilborn, R. and Walters, C.J. (2001) *Quantitative Fisheries Stock
Assessment: Choice, Dynamics and Uncertainty*. Kluwer Academic
Publishers, Boston. (Cited in Froese, 2006.)
