---
title: "Length at maturity analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Length at maturity analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(ggplot2)
```

The `maturity()` function fits a logistic regression of binary maturity (0 = immature, 1 = mature) as a function of a continuous predictor such as length or age:

$$P(\text{mature}) = \frac{1}{1 + e^{-(a + b \cdot L)}}$$

Key outputs are $L_{50}$ and $L_{95}$ — the lengths (or ages) at which 50% and 95% of individuals are expected to be mature — with bootstrap confidence intervals.

## Basic usage

```{r}
library(mustelus)
data(spottail)

lm50 <- maturity(maturity_stage, length, data = spottail, times = 200)
summary(lm50)
```

## Grouping by sex

```{r}
lm50_sex <- maturity(maturity_stage, length, sex, data = spottail, times = 200)
summary(lm50_sex)
```

## Plotting

`plot()` returns a named list of ggplot objects. The default shows binned observed proportions scaled by sample size.

```{r}
p <- plot(lm50_sex)
p$f + xlab("Total length (mm)") + ylab("Proportion mature")
```

### Plot styles

Three styles for raw data display are available:

```{r}
# Proportions (default) — point size scaled by n per bin
plot(lm50, raw_data = "proportions")$Unspecified +
  xlab("Total length (mm)") + ylab("Proportion mature")
```

```{r}
# Point — same bins, uniform point size
plot(lm50, raw_data = "point")$Unspecified +
  xlab("Total length (mm)") + ylab("Proportion mature")
```

```{r}
# Rug — mature individuals on top, immature on bottom
plot(lm50, raw_data = "rug")$Unspecified +
  xlab("Total length (mm)") + ylab("Proportion mature")
```

### Adjusting bin width

The `binwidth` argument controls the width of bins in data units (default is range/10):

```{r}
plot(lm50, binwidth = 50)$Unspecified +
  xlab("Total length (mm)") + ylab("Proportion mature")
```

## Using age as the predictor

`maturity()` works with any continuous predictor — here using age rather than length:

```{r}
am50 <- maturity(maturity_stage, age_agree, data = spottail, times = 200)
summary(am50)
```

```{r}
plot(am50, binwidth = 1)$Unspecified +
  xlab("Age (years)") + ylab("Proportion mature")
```

## Output structure

Like `len_weight()`, the result is a nested tibble with named list columns per group:

```{r}
# Bootstrap L50 estimates for all individuals
quantile(lm50_sex$mods[["f"]]$fitted.values)

# Predicted maturity curve for females
head(lm50_sex$preds[["f"]])
```
